Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-04-18 doc/thesis/intro.tex
Panicked 1AM writing

\textcite{poon2013mapping} developed an alternative pairwise measure which
applies the concept of a \defn{kernel function} to phylogenies. Kernel
functions, originally developed for \glspl{SVM}~\autocite{burges1998tutorial}, 
compare objects by mapping them into a feature space of very high, possibly
even infinite, dimension.

Another tree distance measure is the \defn{phylogenetic kernel}, or ``tree
kernel'' developed by \textcite{poon2013mapping}. As opposed to the \gls{nltt},
the tree kernel is maximized when the two trees being compared are the same.
The basis of the tree kernel is the kernel trick originally developed for
\glspl{SVM}~\autocite{burges1998tutorial}. The idea of the kernel trick is to
compare objects by mapping them into a feature space of very high, possibly
even infinite, dimension. The similarity between objects is taken to be their
dot product in the feature space. It is called a ``trick'' because this dot
product is computed using a \defn{kernel function} without explicitly mapping
the objects to the feature space, which would be computationally prohibitive.
In the case of the tree kernel, the feature space is the space of all possible
\defn{subset trees}, which are subtrees that do not necessarily extend all the
way to the tips. The subset-tree kernel was originally developed for comparing
parse trees in natural language processing~\autocite{collins2002new} and did
not incorporate branch length information. The version developed by
\textcite{poon2013mapping} includes a radial basis function to compare the
differences in branch lengths, thus incorporating both the trees' topologies
and their branch lengths in a single similarity score. The tree kernel was
later shown to be highly effective in differentiating trees simulated under a
compartmental model with two risk groups of varying contact
rates~\autocite{poon2015phylodynamic}. In that paper,
\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
in \gls{ABC} (see \cref{sec:abc}), to fit epidemiological models to observed
trees. This is an example of kernel-ABC~\autocite{nakagome2013kernel}, which
will be discussed further in \cref{sec:abc}.

@@

Contact networks are a challenging type of data to collect, requiring extensive
epidemiological investigation in the form of contact
tracing~\autocite{morris1993epidemiology, welch2011statistical,
keeling2005networks}. Therefore, it has been necessary to explore less
resource-intensive alternatives which still contain information about
population structure. For instance, it is possible to obtain limited
information about the contact network by individual interviews without contact
tracing. Variables which can be estimated in this fashion are referred to as
\defn{node-level} measures~\autocite{wasserman1994social}. One of the most
well-studied of these is the degree distribution, which can be estimated by
simply asking each person how many contacts they had in some interval of
time~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk}.

@@

There has been some contention of the idea that contact networks are
scale-free. \textcite{handcock2004likelihood} fit several stochastic models of
partner formation to empirical degree distributions derived from population
surveys of sexual behaviour. They found that a negative binomial distribution,
rather than a power law, was the best fit to five out of six datasets, although
the difference in goodness of fit was extremely small in four out of these
five. \textcite{bansal2007individual} found that an exponential distribution,
rather than a power law, was the best fit to degree distributions of six social
or sexual networks.

@@

The contact network underlying an epidemic constrains the shape of the
transmission network, which in turn determines the topology of the transmission
tree relating the infected hosts (\cref{fig:contactnet}). The index case who
introduces the epidemic into the network becomes the root of the tree. Each
time a transmission occurs, the lineage corresponding to the donor host in the
tree splits into two, representing the recipient lineage and the continuation
of the donor lineage. \Cref{fig:contactnet} illustrates this correspondence.
It's important to note that, although the order and timing of transmissions
determines the tree topology uniquely, the converse does not hold. That is, for
any given topology, there are in general many transmission networks which would
lead to that topology. In other words, it impossible to distinguish who
transmitted to whom from a transmission tree alone~\autocite{bernard2007hiv}.

A number of studies have made progress in quantifying the relationship between
contact networks and transmission trees. \textcite{o2010contact} simulated
epidemics over networks with four types of degree distribution. They then
estimated the Bayesian skyride~\autocite{minin2008smooth} population size
trajectory in two ways: from the phylogeny, using \gls{MCMC}; and from the
incidence and prevalence trajectories, using the method developed by 
\textcite{volz2009phylodynamics}. They found that the concordance between
the two skyrides, as well as the relationship between the skyride and
prevalence curve, was qualitatively different for each degree distribution.
\textcite{leventhal2012inferring} investigated the relationship between
transmission tree imbalance and several epidemic parameters under four contact
network models, and found that these relationships varied considerably
depending on which model was being considered. \textcite{welch2011network}
simulated transmission trees over networks with varying degrees of community
structure. They found that transmission trees simulated under networks with low
clustering could not generally be distinguished from those simulated under
highly clustered networks, and concluded that contact network clusters do not
affect transmission tree shape. However, more recently,
\textcite{villandre2016assessment} investigated the correspondence between
contact network clusters and transmission tree clusters, and found a moderate
correspondence between the two.

In summary, studies in this group have demonstrated that network structure
profoundly influences tree shape, but have not attempted to quantitatively
infer network parameters from observed trees.

%\section{Approximate Bayesian computation}

%Consider a model $M$, with parameters $\theta$, which we wish to fit to some
%observed data $D$. By ``fit'', we often mean that we want to find particular
%values $\hat{\theta}$ for the parameters which optimize the likelihood of our
%data given those parameters and the model,
%\[
%  \hat{\theta} = \argmax_\theta \Pr(D | M, \theta).
%\]
%This $\hat{\theta}$ is the \defn{maximum likelihood} parameter estimate. Note
%that we are employing a common abuse of notation here, where $\Pr(\cdots)$ is
%being unsed to refer to a probabilty \emph{density} rather than a true
%probability. Alternative to maximum likelihood, we may be interested less in a
%point estimate and more in the posterior distribution of possible values of
%$\theta$ given our data, $\Pr(\theta \mid M, D)$. This will be expanded upon
%below, but for the moment, for illustrative purposes, we restrict our attention
%to the maximum likelihood problem.
%
%If the model we are fitting is sufficiently simple, it may be possible to
%calculate $\hat{\theta}$ directly, using calculus. Most models do not admit
%analytic maximum likeilhood solutions, but if the likelihood any set of
%parameters can be calculated up to a normalizing constant, then
%${\Pr(D \mid M,\theta)}$ can be optimized numerically. A wide range of
%optimization strategies exist, the choice of which to use depending on the
%complexity of the model and whether or not we have access to the gradients of
%the likelihood function with respect to each of the parameters. The majority of
%modelling problems fit into this category, and numerical optimization is
%well-developed and extremely widely used.
%
%However, there are some cases, often when the observed data is of a complex
%type, that explicitly calculating the likelihood of some observed data is
%impossible, even up to a normalizing constant. For example, suppose that we
%want to model a chess player's behaviour. We will set up a simple one-parameter
%model which describes the chess playing process.  The parameter, $a \in [0, 1]$
%indicates the player's eagerness to remove his opponent's pieces from the
%board. We can write down an algorithm for the player's behaviour under such a
%model.
%
%% TODO: don't break this over the page
%
%\begin{algorithmic}
%  \While{the game is not over}
%    \If{I can capture an opponent's piece and $\Uniform(0, 1) < a$}
%      \State{capture the piece}
%    \Else
%      \State{make any other move at random}
%    \EndIf
%  \EndWhile
%\end{algorithmic}
%
%Suppose the observed data are the ending configurations of the board, after the
%player has concluded a game against an opponent with a known value of $a$. The
%model we have designed is very simple, but it is not obvious how to calculate
%the likelihood of a particular ending configuration. Indeed, it seems that the
%only way is to enumerate every possible path the game could have taken, and
%tabulate the ending configurations of each. Clearly, this is infeasible.
%Approximate Bayesian computation was designed for situations like these, where
%exact likelihoods are not available, perhaps due to the model involving an
%algorithm or generative process.

@@

To fully describe \gls{SMC}, we will introduce some notation and terminology.
For a sequence $x_1, \ldots, x_d$, we will write $\vec{x_i}$ to mean the
partial sequence $x_1, \ldots, x_i$.

A \defn{Markov kernel} is the continuous analogue of the transition matrix in a
finite-state Markov model. For some spaces $X$ and $Y$, $K \maps X \times Y \to
[0, 1]$ such that
\begin{align}
    \label{eq:mk}
    \int_Y K(x, y) \d y = 1
\end{align}
for all $x \in X$. This is an ``operational'' definition of Markov kernel which
will be suitable for our purposes. A more precise definition can be found in
\textbf{TODO}. Note that a Markov kernels have nothing to do with the kernel
functions defined in \cref{subsec:treeshape}, other than sharing a name.
Unfortunately, the word ``kernel'' is ubiquitous in mathematics and refers to
many unrelated concepts.

@@

\gls{SIS} applies \gls{IS} to the problem of sampling from a distribution $\pi$ 
on an high-dimensional space, say $\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The
objective is to build up an importance distribution $\eta$ for $\pi$
sequentially. Let $\vec{x_i} = (x_1, \ldots, x_i)$, for $1 \leq i \leq d$. By
the general product rule,
\[
  \pi(\vec{x}) 
  = \pi(x_1) \pi(x_2 \mid x_1) \cdots
    \pi(x_{d-1} \mid \vec{x_{d-2}}) \pi(x_d \mid \vec{x_{d-1}}).
\]
This decomposition is natural in many contexts, particularly for on-line
estimation. For example, in a stateful model like an \gls{HMM}, $x_i$ may
represent the state at time $i$, with $\pi(\vec{x})$ being the posterior
distribution over possible paths. The importance distribution $\eta$ for $\pi$
will be constructed using a similar decomposition,
\[
  \eta(\vec{x}) 
  = \eta(x_1) \eta(x_2 \mid x_1) \cdots
    \eta(x_{d-1} \mid \vec{x_{d-2}}) \eta(x_d \mid \vec{x_{d-1}}).
\]
The importance weights for $\eta$ with respect to $\pi$ are then defined
recursively as
\begin{align}
  \label{eq:sisw}
  w(\vec{x_i}) = \frac{\pi(\vec{x_i})}{\eta(\vec{x_i})}
  = \frac{\pi(x_i \mid \vec{x_{i-1}})\pi(\vec{x_{i-1}})}
         {\eta(x_i \mid \vec{x_{i-1}})\eta(\vec{x_{i-1}})}
  = \frac{\pi(x_i \mid \vec{x_{i-1}})}
         {\eta(x_i \mid \vec{x_{i-1}})}\cdot w(\vec{x_{i-1}}).
\end{align}
Thus, we can choose $\eta(x_i \mid \vec{x_{i-1}})$ such that the variance of
the importance weights is as small as possible at every step, eventually
arriving at a full importance distribution. The process for choosing $\eta$
will differ according to the problem at hand. One potential choice for
$\eta(x_i \mid \vec{x_{i-1}})$ is simply $\pi(x_i \mid \vec{x_{i-1}})$, if it
is possible to compute. For example, in the case of an \gls{HMM}, $\pi(x_i \mid
\vec{x_{i-1}})$ is simply the transition probability from $x_{i-1}$ to $x_i$.
In a Bayesian setting, $\eta$ may be taken to be the prior distribution. In
some applications, it is possible to compute or approximate the form of $\eta$
which minimizes the variance of the weights. The resulting importance
distribution is known as the \defn{optimal
kernel}~\autocite{cappe2007overview}. Choosing each element of the
decomposition of $\eta$ is a problem-specific task (see \textit{e.g.}
\autocite{smith2013sequential,liu2008monte} for many examples).

The recursive definition \cref{eq:sisw} suggests an algorithm for obtaining a
sample from $\pi$. We begin with a sample of $n$ ``particles'' which have been
sampled from the importance distribution $\eta(x_0)$ for $\pi(x_0)$. The
particles are updated and reweighted $d$ times, corresponding to the $d$
elements of the decomposition of $\pi$. At the $i$th step, each particle is
extended to include $x_i$ drawn from the chosen $\eta(x_i \mid \vec{x_{i-1}})$,
and the importance weights are recalculated and normalized.

Of course, $\eta$ is merely an approximation to $\pi$, and may be a fairly poor
one depending on the application. Try as we might to keep the variances of the
weights low, the cumulative errors at each sequential step tend to push many of
the weights to very low values. This results in a poor approximation to $\pi$,
since only a few particles retain high importance weights after all $d$
sequential steps. To mitigate this problem, we periodically apply a resampling
step when the variance in the importance weights becomes too high. Several
different criteria have been proposed for when to resample, but we focus here
on the one described by \textcite{liu2008monte}, namely the decay of the
\gls{ESS} below a prescribed threshold. The \gls{ESS} of the population of
particles is defined as
\[
  \ESS(w) = \frac{n}{1 + \Var(w)},
\]
where $n$ is the number of particles~\autocite{liu2008monte}. When the
\gls{ESS} drops below the threshold, we resample the particles according to
their weights. This results in the removal of low-weight particles from the
population, and also equalizes all the weights. Various resampling strategies
beyond the basic sampling with replacement have been proposed, but we will not
discuss those here. The full \gls{SIS} algorithm with resampling is given as
\cref{alg:sis}. We use a superscript $^{(k)}$ to refer to the $k$th particle,
and for brevity write $w$ for $w(\vec{x_i})$. The variable $T$ is the threshold
for the \gls{ESS}, below which the population is resampled. It is conventional
to take $T = n/2$.

\begin{algorithm}
  \caption{Sequential importance sampling.}
  \begin{algorithmic}
    \For {$k = 1$ to $n$}
      \State Sample $x_0^{(k)}$ from $\eta(x_0)$
      \Comment{Initialize the $k$th particle}
      \State $w^{(k)} \gets \dfrac{\pi\left(x_0^{(k)}\right)}{\eta\left(x_0^{(k)}\right)}$
    \EndFor
    \For {$i = 1$ to $d$}
      \For {$k = 1$ to $n$}
        \State Sample $x_i^{(k)}$ from $\eta\left(x_i \mid \vec{x_{i-1}^{(k)}}\right)$
        \Comment Extend the $k$th particle
        \State $w^{(k)} \gets \dfrac{\pi\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)}{\eta\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)} \cdot w^{(k)}$
      \EndFor
      \State Normalize the weights so that $\sum w = 1$
      \If {$\ESS(w) < T$}
        \State Resample the particles according to $w$
        \For {$k = 1$ to $n$}
          \State $w^{(k)} \gets 1/n$
        \EndFor
      \EndIf
    \EndFor
    \State Sample $n$ particles with probabilities $w$
  \end{algorithmic}
  \label{alg:sis}
\end{algorithm}

>>>>

\textcite{poon2013mapping} developed an alternative pairwise measure which
applies the concept of a \defn{kernel function} to phylogenies. Kernel
functions, originally developed for \glspl{SVM}~\autocite{burges1998tutorial}, 
compare objects in a space $\X$ by mapping them into a feature space $\F$ of
high or infinite dimension via a function $\varphi$. The similarity between the
objects is defined as
\[  
  K(x, x') = <\varphi(x), \varphi(x')>,
\]
that is, the inner product of the objects' representations in the feature
space. Computing $\varphi(x)$ may be computationally prohibitive due to the
dimension of $\F$. The utility of a kernel function $K$ is that it is
constructed in such a way that it can compute the inner product without
explicitly computing $\varphi(x)$. The kernel function developed in
\autocite{poon2013mapping} will henceforth be referred to as the \defn{tree
kernel}. The feature space into which trees are mapped is the space of all
possible \defn{subset trees}, which are subtrees that do not necessarily extend
all the way to the tips. The subset-tree kernel was originally developed for
comparing parse trees in natural language processing~\autocite{collins2002new}
and did not incorporate branch length information. The version developed by
\textcite{poon2013mapping} includes a radial basis function to compare the
differences in branch lengths, thus incorporating both the trees' topologies
and their branch lengths in a single similarity score.

%The tree kernel was later shown to be highly effective in differentiating trees
%simulated under a compartmental model with two risk groups of varying contact
%rates~\autocite{poon2015phylodynamic}. In that paper,
%\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
%in \gls{ABC} (see \cref{sec:abc}), to fit epidemiological models to observed
%trees. This is an example of kernel-ABC~\autocite{nakagome2013kernel}, which
%will be discussed further in \cref{sec:abc}.

@@

Contact networks are a challenging type of data to collect, requiring extensive
epidemiological investigation in the form of contact
tracing~\autocite{morris1993epidemiology, welch2011statistical,
keeling2005networks, eames2015six}. Therefore, it has been necessary to explore
less resource-intensive alternatives which still contain information about
population structure. For instance, it is possible to obtain limited
information about the contact network by individual interviews without contact
tracing. Variables which can be estimated in this fashion are referred to as
\defn{node-level} measures~\autocite{wasserman1994social}. One of the most
well-studied of these is the degree distribution, which can be estimated by
simply asking each person how many contacts they had in some interval of
time~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk}.

@@

There has been some contention of the idea that contact networks are
scale-free. \textcite{handcock2004likelihood} fit several stochastic models of
partner formation to empirical degree distributions derived from population
surveys of sexual behaviour. They found that a negative binomial distribution,
rather than a power law, was the best fit to five out of six datasets, although
the difference in goodness of fit was extremely small in four out of these
five. \textcite{bansal2007individual} found that an exponential distribution,
rather than a power law, was the best fit to degree distributions of six social
and sexual networks.

@@

The contact network underlying an epidemic constrains the shape of the
transmission network, which in turn determines the topology of the transmission
tree relating the infected hosts (\cref{fig:contactnet}). The index case who
introduces the epidemic into the network becomes the root of the tree. Each
time a transmission occurs, the lineage corresponding to the donor host in the
tree splits into two, representing the recipient lineage and the continuation
of the donor lineage. \Cref{fig:contactnet} illustrates this correspondence.
It must be emphasized that, although the order and timing of transmissions
determines the tree topology uniquely, the converse does not hold. That is, for
any given topology, there are in general many transmission networks which would
lead to that topology. In other words, it impossible to distinguish who
transmitted to whom from a transmission tree alone~\autocite{bernard2007hiv}.

A number of studies have made progress in quantifying the relationship between
contact networks and transmission trees. \textcite{o2010contact} simulated
epidemics over networks with four types of degree distribution. They then
estimated the Bayesian skyride~\autocite{minin2008smooth} population size
trajectory in two ways: from the phylogeny, using \gls{MCMC}; and from the
incidence and prevalence trajectories, using the method developed by
\textcite{volz2009phylodynamics}. The concordance between the two skyrides, as
well as the relationship between the skyride and prevalence curve, was
qualitatively different for each degree distribution.
\textcite{leventhal2012inferring} investigated the relationship between
transmission tree imbalance and several epidemic parameters under four contact
network models, and found that these relationships varied considerably
depending on which model was being considered. \textcite{welch2011network}
simulated transmission trees over networks with varying degrees of community
structure. They found that transmission trees simulated under networks with low
clustering could not generally be distinguished from those simulated under
highly clustered networks, and concluded that contact network clusters do not
affect transmission tree shape. However, more recently,
\textcite{villandre2016assessment} investigated the correspondence between
contact network clusters and transmission tree clusters, and found a moderate
correspondence between the two.

@@

To fully describe \gls{SMC}, we will introduce some notation and terminology.
The definitions of these terms will become clearer as they are used. For a
sequence $x_1, \ldots, x_d$, we will write $\vec{x_i}$ to mean the partial
sequence $x_1, \ldots, x_i$. The subscript $^{(k)}$ will be used to indicate
the $k$th particle in a population. To ease the notational burden we will omit
the superscripts and subscripts on the weight functions $w$.

We define a \defn{Markov kernel} as the continuous analogue of the transition
matrix in a finite-state Markov model.  For some spaces $X$ and $Y$, $K \maps X
\times Y \to [0, 1]$ such that
\begin{align}
    \label{eq:mk}
    \int_Y K(x, y) \d y = 1
\end{align}
for all $x \in X$. This is an ``operational'' definition of Markov kernel which
will be suitable for our purposes. A more precise definition can be found in
\textbf{TODO}. Note that a Markov kernels have nothing to do with the kernel
functions defined in \cref{subsec:treeshape}, other than sharing a name (the
word ``kernel'' is ubiquitous in mathematics).

@@

\gls{SIS} applies \gls{IS} to the problem of sampling from a distribution $\pi$ 
on an high-dimensional space, say $\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The
objective is to build up an importance distribution $\eta$ for $\pi$
sequentially. By the general product rule, $\pi(\vec{x})$ can be decomposed as
\[
  \pi(\vec{x}) 
  = \pi(x_1) \pi(x_2 \mid x_1) \cdots
    \pi(x_{d-1} \mid \vec{x_{d-2}}) \pi(x_d \mid \vec{x_{d-1}}).
\]
This decomposition is natural in many contexts, particularly for on-line
estimation. For example, in a stateful model like an \gls{HMM}, $x_i$ may
represent the state at time $i$, with $\pi(\vec{x})$ being the posterior
distribution over possible paths. The importance distribution $\eta$ for $\pi$
will be constructed using a similar decomposition,
\[
  \eta(\vec{x}) 
  = \eta(x_1) \eta(x_2 \mid x_1) \cdots
    \eta(x_{d-1} \mid \vec{x_{d-2}}) \eta(x_d \mid \vec{x_{d-1}}).
\]
The importance weights for $\eta$ can be written recursively as
\begin{align}
  \label{eq:sisw}
  w(\vec{x_i}) = \frac{\pi(\vec{x_i})}{\eta(\vec{x_i})}
  = \frac{\pi(x_i \mid \vec{x_{i-1}})\pi(\vec{x_{i-1}})}
         {\eta(x_i \mid \vec{x_{i-1}})\eta(\vec{x_{i-1}})}
  = \frac{\pi(x_i \mid \vec{x_{i-1}})}
         {\eta(x_i \mid \vec{x_{i-1}})}\cdot w(\vec{x_{i-1}}).
\end{align}
Thus, we can choose $\eta(x_i \mid \vec{x_{i-1}})$ such that the variance of
the importance weights is as small as possible at every step, eventually
arriving at a full importance distribution. This choice is made on a
problem-specific basis, taking any available information about $\pi(x_i \mid
\vec{x_{i-1}})$ into account (see \textit{e.g.}
\autocite{smith2013sequential,liu2008monte} for many examples).  One potential
choice for $\eta(x_i \mid \vec{x_{i-1}})$ is simply $\pi(x_i \mid
\vec{x_{i-1}})$, if it is possible to compute. In a Bayesian setting, the
prior distribution may be used. The exact form of $\eta(x_i, \vec{x_{i-1}})$
which minimizes the variance of the weights is called the \defn{optimal
kernel}~\autocite{cappe2007overview}, the name deriving from the fact that
$k(x_i, \vec{x_{i-1}}) = \eta(x_i, \vec{x_{i-1}})$ is a Markov kernel. In some
applications, it is possible to approximate the optimal kernel or even compute
it explicitly.

The recursive definition \cref{eq:sisw} suggests an algorithm for obtaining a
sample from $\pi$ (\cref{alg:sis}).  We begin with $n$ ``particles'' which have
been sampled from the importance distribution $\eta(x_0)$ for $\pi(x_0)$. The
particles are updated and reweighted $d$ times, corresponding to the $d$
elements of the decomposition of $\pi$. At the $i$th step, each particle is
extended to include $x_i$ drawn according to the chosen $\eta(x_i \mid
\vec{x_{i-1}})$, and the importance weights are recalculated and normalized.

\begin{algorithm}
  \caption{Sequential importance sampling.}
  \begin{algorithmic}
    \For {$k = 1$ to $n$}
      \State Sample $x_1^{(k)}$ from $\eta(x_1)$
      \Comment{Initialize the $k$th particle}
      \State $w^{(k)} \gets \dfrac{\pi\left(x_1^{(k)}\right)}{\eta\left(x_1^{(k)}\right)}$
    \EndFor
    \For {$i = 2$ to $d$}
      \For {$k = 1$ to $n$}
        \State Sample $x_i^{(k)}$ from $\eta\left(x_i \mid \vec{x_{i-1}^{(k)}}\right)$
        \Comment Extend the $k$th particle
        \State $w(\vec{x_i}^{(k)}) \gets \dfrac{\pi\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)}{\eta\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)} \cdot w(\vec{x_{i-1}}^{(k)})$
      \EndFor
      \State Normalize the weights so that $\sum w = 1$
    \EndFor
    \State Sample $n$ particles with probabilities $w$
  \end{algorithmic}
  \label{alg:sis}
\end{algorithm}

Of course, $\eta$ is merely an approximation to $\pi$, and may be a fairly poor
one depending on the application. Try as we might to keep the variances of the
weights low, the cumulative errors at each sequential step tend to push many of
the weights to very low values. This results in a poor approximation to $\pi$,
since only a few particles retain high importance weights after all $d$
sequential steps. To mitigate this problem, we periodically apply a resampling
step when the variance in the importance weights becomes too high. Several
different criteria have been proposed for when to resample, but we focus here
on the one described by \textcite{liu2008monte}, namely the decay of the
\gls{ESS} below a prescribed threshold, conventionally $n/2$. The \gls{ESS} of
the population of particles is defined as
\[
  \ESS(w) = \frac{n}{1 + \Var(w)},
\]
where $n$ is the number of particles~\autocite{liu2008monte}. When the
\gls{ESS} drops below the threshold, we resample the particles according to
their weights. This results in the removal of low-weight particles from the
population, and also equalizes all the weights. Various resampling strategies
beyond the basic sampling with replacement have been proposed, but we will not
discuss those here.
